Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-01-30 WholePlotTrajectories.Rmd
Discussion IDH

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the resilience of community taxonomic and functional diversity and composition remains controversial. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed tree community trajectories following a disturbance gradient in a Neotropical forest. We examined community diversity, composition and redundancy trajectories over 30 years, considering both taxonomic and functional trajectories based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
  bookdown::gitbook: default
---

@@

The IDH then assumes both predictable trajectories depending on the disturbance intensity, and, without additional disturbance, the recovery of pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Sheil2003;@Willig2018].
In highly diverse tropical forests, however, empirical validation of the IDH remains debated [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hugues2007].

Still, empirical evidence often relied on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], and this might give limited or misleading information on forests recovery and functioning. 
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

@@

Community taxonomic richness and evenness trajectories both proved to depend on the intensity of disturbance and to markedly change above an intensity threshold, thus supporting the IDH.
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below an intensity of 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community.
The recruitment of pionners, infrequent or absent before disturbance, then increased the community taxonomic richness positively correlated with the disturbance intensity [@Martin2015; @Chaudhary2016].
Besides, the pioneers became increasingly dominant and temporarily balanced the usual hyper-dominance of a few species [@Baraloto2012a]. 
The taxonomic evenness then followed hump-shaped trajectoriesas already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998], with first an increase positively correlated to the disturbance intensity, and then an return towards initial values.
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The taxonomic richness trajectories following intense disturbance then decreased, all the more so that the disturbance intensity was high [@Molino2001].
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.
The intensity threshold then marked two distinct post-disturbance trajectories, and a change in community response to disturbance. 
In accordance with the IDH, an intense disturbance below the intensity threshold would then maximise community richness and evenness.
Such intermediate disturbance would increase the availability of resources, creating opportunities for pioneers, while allowing a high diversity among surviving trees and the maintenance of shade-tolerant species [@Bongers2009].

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the resilience of community taxonomic and functional diversity and composition remains controversial. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed tree community trajectories following a disturbance gradient in a Neotropical forest. We examined community diversity, composition and redundancy trajectories over 30 years, considering both taxonomic and functional trajectories based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::word_document2: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
---

@@

The IDH then assumes both predictable trajectories depending on the disturbance intensity, and, without additional disturbance, the recovery of pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Sheil2003;@Willig2018].
In highly diverse tropical forests, however, the validation of the IDH remains controversial [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hugues2007].
Empirical evidence generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which might give limited or misleading information on community disturbance response. 
The IDH pattern of diversity peak under an intermediate disturbance regime can indeed result from various ecological processes [@Sheil2003].
To disentangle these various processes and grasp the different facets of community response to disturbance it is crucial to combine a suitable range of approaches [@Shea2004].
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

@@

Community taxonomic richness and evenness trajectories were determined by the disturbance intensity, ranging from a limited and temporary impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold for which community response corresponded to a maximized richness and a resilience of the evenness. 
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees, and trees recruited afterward. 
Below a 20%-25% AGB loss, the trees surviving after disturbance remained numerous enough to maintain the high taxonomic richness of the pre-disturbance community [@Bongers2009].
The recruitment of pionners, infrequent or absent before disturbance, then increased the taxonomic richness all the more so that the disturbance was intense [@Martin2015; @Chaudhary2016].
As these pioneers became more dominant they besides temporarily balanced the usual hyper-dominance of tropical forests, hence temporarily increasing the taxonomic evenness [@Baraloto2012a].
Above the intensity threshold, the taxonomic richness of surviving trees was too low to be offset by the recruitment of pioneers. 
In the Guiana shield indeed, the pool of true pioneers specifically recruited after disturbance is restricted to a few common genera (e.g. *Cecropia* spp., *Vismia* spp.) [@Guitet2018].
The richness trajectories following intense disturbance then decreased, all the more so that the disturbance intensity was high [@Molino2001]. 
At the same time, pioneers became persistently dominant and prevented a return towards the initial evenness and the recovery of hyper-dominant shade-tolerant species.

Observed trajectories supported the IDH in highlighting a threshold disturbance intensity, expressed as percentage of AGB removed, marking a change in post-disturbance trajectories.
Taxonomic trajectories besides highlighted, below the intensity threshold, an intermediate time after when the taxonomic evenness was maximized.
This time corresponded to the intermediate between the generation time of pionneers, first to dominate the post-disturbance recruitment, and the time for competitive exclusion, leading to the dominance of late-successional species.
As already observed in the Guiana shield [@Baraloto2012a] and in Borenan tropical forests [@Cannon1998],
taxonomic evenness followed humped-shaped trajectories with a peak when the recruitment was balanced between pioneers and late-successional species.

### 2019-01-30 WholePlotTrajectories.tex
Discussion IDH

The IDH then assumes both predictable trajectories depending on the
disturbance intensity, and, without additional disturbance, the recovery
of pre-disturbance state defined by the dominance of late-successional
species \citep{Rees2001, Sheil2003, Willig2018}. In highly diverse
tropical forests, however, empirical validation of the IDH remains
debated \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hugues2007}.

Still, empirical evidence often relied on the analysis of a single
diversity metric, often the taxonomic richness \citep{Molino2001}, and
this might give limited or misleading information on forests recovery
and functioning. More ecological-meaningful analysis would first couple
taxonomic richness and evenness, which reveals the changes in species
abundance distribution and hence in the underlying ecological processes
\citep{Chaudhary2016}. Analysis would as well include taxonomic
composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006}. Eventually, including a functional
approach accounting for species biological attributes would be highly
revealing of ecological process involved in community response to
disturbance. The functional trait-based approach indeed directly links
community diversity, composition and redundancy to ecosystem functioning
and environmental constraints \citep{Violle2007b, Baraloto2012a}. In
that respect, a vast literature allowed recognizing major traits related
to species ecology and mediating species performance in a given
environment \citep{Diaz2005}. Specifically in tropical forests, the
functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

@@

Community taxonomic richness and evenness trajectories both proved to
depend on the intensity of disturbance and to markedly change above an
intensity threshold, thus supporting the IDH. The disturbance intensity
determined the balance in the community between pre-disturbance
surviving trees, and trees recruited afterward. Below an intensity of
20\%-25\% AGB loss, the trees surviving after disturbance remained
numerous enough to maintain the high taxonomic richness of the
pre-disturbance community. The recruitment of pionners, infrequent or
absent before disturbance, then increased the community taxonomic
richness positively correlated with the disturbance intensity
\citep{Martin2015, Chaudhary2016}. Besides, the pioneers became
increasingly dominant and temporarily balanced the usual hyper-dominance
of a few species \citep{Baraloto2012a}. The taxonomic evenness then
followed hump-shaped trajectoriesas already observed in the Guiana
shield \citep{Baraloto2012a} and in Borenan tropical forests
\citep{Cannon1998}, with first an increase positively correlated to the
disturbance intensity, and then an return towards initial values. Above
the intensity threshold, the taxonomic richness of surviving trees was
too low to be offset by the recruitment of pioneers. In the Guiana
shield indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
trajectories following intense disturbance then decreased, all the more
so that the disturbance intensity was high \citep{Molino2001}. At the
same time, pioneers became persistently dominant and prevented a return
towards the initial evenness and the recovery of hyper-dominant
shade-tolerant species. The intensity threshold then marked two distinct
post-disturbance trajectories, and a change in community response to
disturbance. In accordance with the IDH, an intense disturbance below
the intensity threshold would then maximise community richness and
evenness. Such intermediate disturbance would increase the availability
of resources, creating opportunities for pioneers, while allowing a high
diversity among surviving trees and the maintenance of shade-tolerant
species \citep{Bongers2009}.

>>>>

The IDH then assumes both predictable trajectories depending on the
disturbance intensity, and, without additional disturbance, the recovery
of pre-disturbance state defined by the dominance of late-successional
species \citep{Rees2001, Sheil2003, Willig2018}. In highly diverse
tropical forests, however, the validation of the IDH remains
controversial \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hugues2007}. Empirical evidence generally rely on the analysis of
a single diversity metric, often the taxonomic richness
\citep{Molino2001}, which might give limited or misleading information
on community disturbance response. The IDH pattern of diversity peak
under an intermediate disturbance regime can indeed result from various
ecological processes \citep{Sheil2003}. To disentangle these various
processes and grasp the different facets of community response to
disturbance it is crucial to combine a suitable range of approaches
\citep{Shea2004}. More ecological-meaningful analysis would first couple
taxonomic richness and evenness, which reveals the changes in species
abundance distribution and hence in the underlying ecological processes
\citep{Chaudhary2016}. Analysis would as well include taxonomic
composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006}. Eventually, including a functional
approach accounting for species biological attributes would be highly
revealing of ecological process involved in community response to
disturbance. The functional trait-based approach indeed directly links
community diversity, composition and redundancy to ecosystem functioning
and environmental constraints \citep{Violle2007b, Baraloto2012a}. In
that respect, a vast literature allowed recognizing major traits related
to species ecology and mediating species performance in a given
environment \citep{Diaz2005}. Specifically in tropical forests, the
functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which community response corresponded to a maximized richness and a
resilience of the evenness. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pionners, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they besides temporarily balanced
the usual hyper-dominance of tropical forests, hence temporarily
increasing the taxonomic evenness \citep{Baraloto2012a}. Above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The richness trajectories
following intense disturbance then decreased, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

Observed trajectories supported the IDH in highlighting a threshold
disturbance intensity, expressed as percentage of AGB removed, marking a
change in post-disturbance trajectories. Taxonomic trajectories besides
highlighted, below the intensity threshold, an intermediate time after
when the taxonomic evenness was maximized. This time corresponded to the
intermediate between the generation time of pionneers, first to dominate
the post-disturbance recruitment, and the time for competitive
exclusion, leading to the dominance of late-successional species. As
already observed in the Guiana shield \citep{Baraloto2012a} and in
Borenan tropical forests \citep{Cannon1998}, taxonomic evenness followed
humped-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species.
